Hierarchical recurrent models reveal latent states of neural activity in \textit{C. elegans}

### 2017-12-21 drafts/elife/draft.tex
adding results on discrete states

>>>>

Having learned a mapping from latent state space to observed activity,
we can now reason about the dynamics of these continuous latent
states. Fig.~\ref{fig:syllables}a plots the trajectory of the first
worm's neural activity in the first three dimensions of the continuous
state space, color coded by the discrete regime inferred under the
SLDS.  We see that different regimes correspond to distinct loops
through latent space.  Compare this with Fig. 4b
of~\citet{kato2015global}---they achieved a similar result by manually
segmenting smoothed trajectories in the space spanned by the first
three principal components.  In contrast, our probabilistic framework
is automatic, operates on ten-dimensional latent states, and requires
no additional smoothing.

Moreover, we can use the likelihood of held-out test data to determine
the number of discrete regimes and assess the value of the robust and
recurrent model extensions, as shown in Fig.~\ref{fig:syllables}b.
While the non-robust models with Gaussian dynamics noise achieve
higher training likelihood, the robust models generalize better to
test data.  Moreover, the recurrent models provide further improvement
over the standard Markovian models.  We find that test likelihood is
maximized with eight latent states, which matches the number manually
chosen by~\citet{kato2015global}.

Not only does the SLDS find the same number of states, it also finds a
close correspondence between the manually labeled and inferred states.
Fig.~\ref{fig:syllables}c shows ten-dimensional latent states and the
manual and inferred segmentations for a 5.5 minute section of data.
The manual and inferred states often change at similar times. We also
see instances where there appears to be a change in continuous state
dynamics that are identified by the SLDS but missed in the manual
segmentation (e.g. around $t=700$ seconds).  Fig.~\ref{fig:syllables}d
quantifies the overlap in states for the first worm and for all worms.
Many inferred states are in one-to-one correspondence with manually
labeled states, and the remainder tend to be split between two manual
states.  This confirms that the SLDS is a useful tool for automatically
segmenting brain data.

Beyond simply automating a manual process, the SLDS also provides a
quantitative description of the different states.  In particular, each
state is associated with a set of linear dynamics.
Fig.~\ref{fig:syllables}e shows the canonical dynamics associated with
each of the eight discrete states. The arrows in these vector fields
show where the continuous state is mapped to under the corresponding
discrete state (for the first three latent dimensions).  Arrows are
only drawn at those locations where the discrete state was actually
employed.  We see that these dynamics capture the rotational patterns
of the various loops.

We use these dynamics to simulate how brain activity would evolve
under the different states. Fig.~\ref{fig:syllables}f shows 200
simulations of state 4 from various starting locations.  Four
simulations are highlighted in particular, as denoted by the outlined
circle, triangle, square, and pentagon.  We then map these
trajectories into the space of neural activity and show how they would
activate the five clusters shown in Fig.~\ref{fig:emissions}.  We see
that state 4 corresponds to a pronounced dip followed by a rise in the
activity of cluster 3, alongside a slow ramp in the activity of
clusters 1 and 2. Referring back to Fig~\ref{fig:emissions}a, we see
that cluster 3 consists of \textsf{ALA} and three symmetric pairs of
neurons (\textsf{ASK}, \textsf{RIV}, and \textsf{SMDV}).  The~\textsf{ASK}
pair are sensory neurons responsible for initiating local search behavior,
the \textsf{RIV} pair are polymodal inter- and motor-neurons that specify
the ventral bias of turns, and the \textsf{SMDV} are motor neurons
that initiate ventral turns. As state 4 is in close correspondence with
the ``ventral turn'' state labeled by~\citet{kato2015global}, it makes
sense that these neurons would be activated.
